---
title: "Present a great story for data science projects"
subtitle: "This is a Slide Template"
author: "Students names (Advisor: Dr. Cohen)"
date: '`r Sys.Date()`'
format:
  revealjs
course: Capstone Projects in Data Science
bibliography: references.bib # file contains bibtex for references
#always_allow_html: true # this allows to get PDF with HTML features
self-contained: true
execute: 
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

::: callout-important
**Remember:** Your goal is to make your audience understand and care
about your findings. By crafting a compelling story, you can effectively
communicate the value of your data science project.

Carefully read this template since it has instructions and tips to
writing!

More information about `revealjs`:
<https://quarto.org/docs/reference/formats/presentations/revealjs.html>
:::

## Introduction  {.smaller}

In financial institutions, evaluating credit risk is essential since it has a direct bearing on lending choices, portfolio performance, and regulatory compliance. The simplicity and interpretability of traditional credit scoring models, including logistic regression, have led to their widespread use. Unfortunately, these models are less successful due to the increasing number and complexity of financial data, as they are unable to capture complicated borrower behavior and non-linear correlations. 
Credit risk prediction has been greatly enhanced by recent developments in machine learning, particularly ensemble techniques like Gradient Boosting Decision Trees. Large, high-dimensional, and unbalanced credit datasets can be handled using the scalable and potent Light Gradient Boosting Machine (LightGBM) model. LightGBM is frequently regarded as a black-box approach, which restricts transparency and erodes confidence among regulators and business users despite its impressive performance.
This study suggests an Explainable Agentic AI framework for credit risk assessment utilizing LightGBM in order to overcome these issues. An intelligent agentic decision layer that can read predictions, produce explanations, and suggest actions is combined with LightGBM and SHAP-based explainability. The system seeks to promote responsible and explicable credit judgments in regulated financial contexts by combining accuracy, transparency, and adaptive decision-making.

## Methods

-   Detail the models or algorithms used.

-   Justify your choices based on the problem and data.



## Data Exploration and Visualization

-   Describe your data sources and collection process.

-   Present initial findings and insights through visualizations.

-   Highlight unexpected patterns or anomalies.

## Data Exploration and Visualization {.smaller}

A study was conducted to determine how...

```{r, warning=FALSE, echo=F, message=FALSE}
# loading packages 
library(tidyverse)
library(knitr)
library(ggthemes)
library(ggrepel)
library(dslabs)
```

```{r, warning=FALSE, echo=F}
# Load Data
#kable(head(murders))

ggplot1 = murders %>% ggplot(mapping = aes(x=population/10^6, y=total)) 

  ggplot1 + geom_point(aes(col=region), size = 4) +
  geom_text_repel(aes(label=abb)) +
  scale_x_log10() +
  scale_y_log10() +
  geom_smooth(formula = "y~x", method=lm,se = F)+
  xlab("Populations in millions (log10 scale)") + 
  ylab("Total number of murders (log10 scale)") +
  ggtitle("US Gun Murders in 2010") +
  scale_color_discrete(name = "Region")+
      theme_bw()
  

```

## Modeling and Results

-   Explain your data preprocessing and cleaning steps.

-   Present your key findings in a clear and concise manner.

-   Use visuals to support your claims.

-   **Tell a story about what the data reveals.**

```{r}

```

## Conclusion

-   Summarize your key findings.

-   Discuss the implications of your results.

## References
